The competing effects of the immune system and EMT on epithelial cancers

### 2019-04-19 Paper/Paper.tex
Small Edits and a TGFB Source

The effects of the immune system on a tumor can be broadly summarized into two modules. The cytotoxic branch of the immune system (such as natural killer (NK) cells and cytotoxic T cells (CTLs) seek out and lyse tumor cells.

@@

The regulatory branch of the immune system (T regulatory cells (Tregs) and other factors), inhibits the effective functioning of the cytotoxic branch\cite{ruffell2010lymphocytes}.

@@

\section{Methods}
We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo2017multiscale}. 
The agents in the model are the tissue cells with tumorigenic potential.
These can either be mutation-free, or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).
If the pathway is mutated, the component is colored red and is marked with `X'.
A single mutation is sufficient for the cell to be considered as a mutant cell.
\par
We model immune cells as continuous variables, appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.
The cytokine TGF-$\beta$ is continuous within the tumor microenvironment (TME), i.e. the system is well-mixed.
Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
In Figure \ref{fig:ModelIntro}A, the EMT labeling is given by the color of the cell.
Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.

@@

\subsubsection{Initial conditions} Every model run starts with a fixed amount of cells, $N_0$.
However, different parameter values will lead to different steady states for the population of tissue cells.
In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.
Thus, the only immune cell present is NKs.
After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
From then on, a cell which proliferates either mutates or increases its probability of mutating later on.

@@

\paragraph{Proliferation}
There are four factors that contribute to the probability that a cell will proliferate.
The first is a base proliferation rate that all cells have, $p$.
Second, if the cell has a mutation in the proliferation pathway ($\delta_P=1$), then the weight for proliferation is proportionally increased by $\Delta_P$.
Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
In total, the weight for proliferation is given by

@@

\subsubsection{Completing the Cell Cycle}
After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
Second, all proliferating cells have a nonzero probability of undergoing a pathway mutation.
If they do, one is randomly chosen among the three pathways and the pathway in that cell mutates.
If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases. 
Finally, the EMT values for each cell is updated.
This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
Each cell receives an amount of TGF-$\beta$ given by

@@

The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.
Then, some randomness is introduced by the independent and identically distributed variables $X_i$where $\sigma$ is a parameter in the model.
If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
If not, then the cell undergoes the reverse process, MET, and its EMT score will decrease.
Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: quantity of TGF-$\beta$ in the system and the current EMT score of the cell.

@@

\begin{equation}\tag{2.9}
\begin{split}
N_\text{CTL}' & = \sigma_{\text{CTL}}N_{\text{MUT}}^* - d_{\text{CTL}}\text{CTL} \\
\text{Treg}' & = \sigma_{\text{Treg}}N_{\text{MUT}}^* \frac{\tau}{1+\tau/K_4}- d_{\text{Treg}}N_\text{Treg}
\end{split}
\end{equation}

@@

\begin{equation}\tag{2.10}
\begin{split}
N_\text{CTL}(k+1) & =  \left (N_\text{CTL}(k)-\sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\\
N_\text{Treg}(k+1) & =  \left (N_\text{Treg}(k)-\frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau}{1+\tau/K_4}\right )\exp(-d_{\text{Treg}}\Delta t)\\
&+ \frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau}{1+\tau/K_4}
\end{split}
\end{equation}

@@

\subsection{Parameter Sensitivity Analysis}
To study parameter sensitivity, we implemented the Morris method, a one-step-at-a-time global sensitivity analysis algorithm.
%The algorithm runs the model numerous successive times, \tcr{changing exactly one parameter each run} \cite{morris1991factorial, sohier2014improvement}.
The algorithm runs the model numerous times by picking several base points and then varying one parameter at a time from each base point and tracking the difference\cite{morris1991factorial, sohier2014improvement}.
Each model run involved 1000 patients and the value returned was the area under the survival curve.
We ran the algorithm by sampling at 30 points in parameter space and then changing each parameter one time at each point.
In the implementation by Sohier, they recommend using at least 10 points\cite{sohier2014improvement}.
\par
To determine which points in parameter space to sample at, we had to create distributions on all parameter values.
To do this, we derived many parameters from existing literature and the rest we made reasoned predictions.
For example, many previous studies have established the parameters of immune population dynamics such as recruitment and death rates\cite{de2014modeling}.
Because our model only works with on the order of $10^2$ tissue cells whereas an actual tumor would have on the order of $10^9$ cells\cite{de2014modeling}, we scaled the parameters so the immune populations would also be proportionately reduced.
That is, for instance, how we ended up with NK cells having a steady state population of 10 cells in the model.
\par
Once we had these parameter values, we created Gaussian distributions centered at these values with large variances and truncated at 0.
Some of the parameters were also truncated to the right because they represented a proportion, such as MIE and MGA, or because we simply wanted to keep the values within a reasonable range, such as the inflammation low and high duration parameters.
The variances were chosen as twice the mean to get a large spread of values.
\par
%The algorithm \tcr{outputs the average of the absolute change of area under the curve for each parameter normalized by the corresponding change in the cdf function for that parameter. [I don't get this: what is y-axis?]}
The algorithm the average of the absolute change of the output, which in our model is the area under the survival curve.
This output, $\mu^*$, is the sensitivity plotted in Fig. \ref{fig:MOAT}.

@@

We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the probability of cancer, and the Time to Cancer. 
Within the cell cycle, cell fate is determined via a set of rules that are influenced by EMT and immune interactions (Fig. \ref{fig:ModelIntro}A). For example, if a cell undergoes EMT, the probability that it will proliferate is reduced; if a cell gains a mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced. The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset). NK cells and CTLs attempt to clear mutated tissue cells, and deactivate upon successfully carrying out their cytotoxic function. T regulatory cells (Tregs) inhibit cytotoxic activity. Tregs also release TGF-$\beta$ which promotes the recruitment of Tregs, as well as increasing the rate of EMT.
\par
The inflammation cycling scheme for a typical {\it in silico} patient consists of alternating high and low regimes (Fig. \ref{fig:ModelIntro}B); the inflammation schemes modeled will be discussed in detail in the next section. For this patient, after the warmup period, cell mutations are observed at a rate low enough that they are cleared by cytotoxic cells before being able to establish a tumor. However, after approximately 1000 cell cycles (750 days), one mutated cell exhibits clonal growth, and a tumor is established. At this point, large numbers of cells from the adaptive immune system (CTLs and Tregs) are being heavily recruited, and a peak in TGF-$\beta$ expression is observed. The number of mesenchymal cells increases initially with increasing TGF-$\beta$, but decreases after the clone takes over the tissue. This shifts back towards an epithelial tissue phenotype occurs even though the expression of TGF-$\beta$ remains high.
After 841 cell cycles, the proportion of mutated cells reaches 50\%: this is modeled as the threshold defining tumorigenesis. Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days. 
After cancer incidence, the number of mutated cells continues to grow rapidly and soon makes up 100\% of the cell population. A peak in mesenchymal phenotypes is observed following cancer incidence, before most transition back to an epithelial state.
\par 
Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of oncogenic mutations.
The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation. Within each of these periods, adaptive immune populations increase or decrease rapidly in accordance with the inflammation state.
\par 
The {\em in silico} patient described here is given for the purpose of illustrating features of the model. Given the multiple sources of noise in the model, in order to quantify patient dynamics and cancer-free survival rates, below we will simulate large cohorts of patients. In Fig. \ref{fig:ModelIntro}C we simulate the survival curve for a cohort of $500$ patients: we see that all the patients survive for approximately 100 cell cycles (75 days). Subsequently, in the approximate range of $T= [100,300]$, the cancer onset rate is roughly constant, and after $600$ cell cycles, no patients remain cancer free.

\subsection{Identification of key regulatory parameters via Morris global sensitivity analysis}\label{SensAnalysis}
To assess the sensitivity of the model parameters and thus identify those that are most important in determining cancer-free survival times, we applied Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods) to the model.
The results of Morris OAT on the 31 model parameters are shown in \ref{fig:MOAT}: a subset of parameters demonstrate much higher levels of sensitivity than others.
The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
This ties Treg cells to all three components of the model.
Since we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for detailed analysis below.
\par
Many parameters in the model change depending on the inflammation state.
For the sake of a naming convention, any parameter which includes ``low'' in the name represents the parameter value when inflammation is low.
For when the inflammation is high, these ``low'' parameters get scaled by parameters which include ``up'' in the name.
The ILD and IHD parameters determine the duration of low and high inflammation, respectively.
A cell with mutated pathways will be affected by the $\Delta_A$, $\Delta_P$, and $\Delta_{\text{IE}}$ parameters, depending on whether the apoptosis, proliferation, or immune evasion pathways are mutated.
The mesenchymal cells are influenced by the MGA and MIE parameters.
Finally, $\sigma$, $k_\text{EMT}$, $\tau_\text{max}$, and $K_3$ all control how cells transition between the epithelial and mesenchymal states.

@@

\begin{figure}
\center
{\includegraphics[width=0.85\textwidth]{Figure3/Figure3.pdf}}
\caption{Effects of mesencymal cell properites on the time to cancer. Trajectories of one patient per cohort from warmup period ([-1000,0]) to 2000 cell cycles, for $\Delta_\text{MIE}$ (A); $\Delta_\text{MGA}$ (B); and $\tau_\text{Treg}$, Treg TGF-$\beta$ (C). 
D. Survival curves corresponding to changes in MIE (A) for a patient cohort of size 1000; shaded region represents the 95\% confidence interval for the evaluated function. 
E. Survival curves for patient cohort corresponding to changes in MGA (B).
F. Survival curves for patient cohort corresponding to changes in Treg TGF-$\beta$ (C). 
%% \color{red} Cox regression or KM test statistics to be computed to demonstrate the significance of this claim.
}
\label{fig:FirstSurvivalCurves}
\end{figure}

When a cell transitions from an epithelial to a mesenchymal state, two phenotypic cell characteristics change: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
Both parameters are defined as proportions, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
MIE is the proportional reduction in the probability that a mutated mesenchymal cell will be subject to immune clearance in a given cycle.
MGA is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
The cytokine TGF-$\beta$ is also involved in the EMT process (by increasing the probability of EMT). All three of these parameters were found to have large effects on the model by the Morris OAT analysis presented in Section \ref{SensAnalysis}.
\par 
As the level of mesenchymal immune evasion (MIE) is increased, the Time to Cancer decreases (Fig. \ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of potentially mutated cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
These results are summarized by Figure \ref{fig:FirstSurvivalCurves}A.

Second, as the rate of mesenchymal growth arrest (MGA) increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Figure \ref{fig:FirstSurvivalCurves}B).
This is not immediately intuitive, since the decreased proliferation rate affects both mutated and non-mutated tissue cells.

@@

We explored the effects of varying the inflammation state of the patient on cancer-free survival, to investigate competing interactions within the TME and their effect on EMT. Patient cohorts were simulated under different inflammation regimes: permanently low inflammation; permanently high inflammation; or variable inflammation. For patients drawn from cohorts in a permanently high inflammatory state, the relationship between mesenchymal parameters and the Time to Cancer is monotonic, i.e. increasing either the rate of mesenchymal immune evasion (MIE) (Fig. \ref{fig:VaryINFL_and_MesPars}A-B) or of mesenchymal growth arrest (MGA)  (Fig. \ref{fig:VaryINFL_and_MesPars}C-D) decreases the Time to Cancer.
However, under regimes with either temporary or permanent periods of low inflammation, different relationships emerge: a local maximum for the Time to Cancer is found with respect to the relative mesenchymal growth rate (MGA $= 0.3$). This is seen both for intermittent or low inflammation (Fig. \ref{fig:VaryINFL_and_MesPars}D). 
\par
These striking differences in the mean Time to Cancer -- extended by up to one year by fine-grained optimization of mesenchymal growth arrest -- have clear therapeutic implications. 
Our model predicts that reducing the inflammatory state of the TME in three weeks out of every nine would be provide means to prolong cancer-free survival in combination with mesenchymal cell therapies to make the control of mesenchymal proliferation an effective anti-cancer strategy.
\par
In contrast, when MIE is varied for different inflammation cycling schemes, regardless of the cycling scheme, increasing the MIE leads to decreases in the Time to Cancer (worse survival probabilities), although in the case where the inflammatory state is permanently high, MIE has little effect on the Time to Cancer. Thus under any inflammation regime with periods of low inflammation, any reductions in the rate of mesenchymal immune evasion will lead to improved patient outcomes.

>>>>

The effects of the immune system on a tumor can be broadly summarized into two modules.
The cytotoxic branch of the immune system, such as natural killer cells (NKs) and cytotoxic T cells (CTLs), seek out and lyse tumor cells.

@@

The regulatory branch of the immune system, such as T regulatory cells (Tregs) and other factors, inhibits the effective functioning of the cytotoxic branch\cite{ruffell2010lymphocytes}.

@@

\section{Methods}
We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo2017multiscale}. 
The agents in the model are the tissue cells with tumorigenic potential.
These can either be mutation-free or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).
If the pathway is mutated, the component is colored red and is marked with `X'.
A single mutation is sufficient for the cell to be considered as a mutant cell.
\par
We model immune cells as continuous variables, appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.
The cytokine TGF-$\beta$ is continuous within the TME, i.e. the system is well-mixed.
Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
In Figure \ref{fig:ModelIntro}A, the EMT labeling is given by the color of the cell.
Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.

@@

\subsubsection{Initial conditions} Every model run starts with a fixed amount of cells, $N_0$.
However, different parameter values will lead to different steady states for the population of tissue cells.
In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.
Thus, the only immune cell present is NKs.
After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
From then on, a cell which proliferates also either mutates or increases its probability of mutating later on.

@@

\paragraph{Proliferation}
There are four factors that contribute to the weight of a cell to proliferate.
The first is a base proliferation rate that all cells have, $p$.
Second, if the cell has a mutation in the proliferation pathway ($\delta_P=1$), then the weight for proliferation is proportionally increased by $\Delta_P$.
Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
In total, the weight for proliferation is given by

@@

\subsubsection{Completing the Cell Cycle}
After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
Second, all proliferating cells have a cell-specific probability of undergoing a pathway mutation.
If they do, one is randomly chosen among the three pathways and the pathway in that cell mutates.
If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases. 
Finally, the EMT values for each cell is updated.
This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
Each cell receives an amount of TGF-$\beta$ given by

@@

The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.
Then, some randomness is introduced by the independent and identically distributed variables $X_i$ where $\sigma$ is a parameter in the model.
If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
If not, then the cell undergoes the reverse process, MET, and its EMT score will decrease.
Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: concentration of TGF-$\beta$ in the system and the current EMT score of the cell.

@@

\begin{equation}\tag{2.9}
\begin{split}
N_\text{CTL}' & = \sigma_{\text{CTL}}N_{\text{MUT}}^* - d_{\text{CTL}}N_\text{CTL} \\
\text{Treg}' & = \sigma_{\text{Treg}}N_{\text{MUT}}^* \frac{\tau}{1+\tau/K_4}- d_{\text{Treg}}N_\text{Treg}
\end{split}
\end{equation}

@@

\begin{equation}\tag{2.10}
\begin{split}
N_\text{CTL}(k+1) & =  \left (N_\text{CTL}(k)-\sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\right )\exp(- d_{\text{CTL}}\Delta t) + \sigma_{\text{CTL}}N_{\text{MUT}}^*(k)/d_{\text{CTL}}\\
N_\text{Treg}(k+1) & =  \left (N_\text{Treg}(k)-\frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau(k)}{1+\tau(k)/K_4}\right )\exp(-d_{\text{Treg}}\Delta t)\\
&+ \frac{\sigma_{\text{Treg}}N_{\text{MUT}}^*(k)}{d_{\text{Treg}}} \frac{\tau(k)}{1+\tau(k)/K_4}
\end{split}
\end{equation}

@@

\subsection{Parameter Sensitivity Analysis}
To study parameter sensitivity, we implemented the Morris method, a one-step-at-a-time global sensitivity analysis algorithm.
%The algorithm runs the model numerous successive times, \tcr{changing exactly one parameter each run} \cite{morris1991factorial, sohier2014improvement}.
The algorithm runs the model numerous times by picking several base points and then varying one parameter at a time from each base point and tracking the difference\cite{morris1991factorial, sohier2014improvement}.
Each model run involved 1000 patients and the value returned was the area under the survival curve.
We ran the algorithm by sampling at 30 points in parameter space and then changing each parameter one time at each point.
In the implementation by Sohier, they recommend using at least 10 points\cite{sohier2014improvement}.
\par
To determine which points in parameter space to sample at, we had to create distributions on all parameter values.
To do this, we derived many parameters from existing literature and the rest we made reasoned predictions.
For example, many previous studies have established the parameters of immune population dynamics such as recruitment and death rates\cite{de2014modeling}.
Because our model only works with on the order of $10^2$ tissue cells whereas an actual tumor would have on the order of $10^9$ cells\cite{de2014modeling}, we scaled the parameters so the immune populations would also be proportionately reduced.
That is, for instance, how we ended up with NK cells having a steady state population of 10 cells in the model.
\par
Once we had these parameter values, we created Gaussian distributions centered at these values with large variances and truncated at 0.
Some of the parameters were also truncated to the right because they represented a proportion, such as MIE and MGA, or because we simply wanted to keep the values within a reasonable range, such as the inflammation low and high duration parameters.
The variances were chosen as twice the mean to get a large spread of values.
\par
%The algorithm \tcr{outputs the average of the absolute change of area under the curve for each parameter normalized by the corresponding change in the cdf function for that parameter. [I don't get this: what is y-axis?]}
The algorithm computes the average of the absolute change of the output, which in our model is the area under the survival curve.
This output out the algorithm, $\mu^*$, is the sensitivity plotted in Fig. \ref{fig:MOAT}.

@@

We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the probability of cancer, and the Time to Cancer. 
Within the cell cycle, cell fate is determined via a set of rules that are influenced by EMT and immune interactions (Fig. \ref{fig:ModelIntro}A). For example, if a cell undergoes EMT, the probability that it will proliferate is reduced; if a cell gains a mutation in the apoptosis pathway, the probability that it undergoes apoptosis is greatly reduced. The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset). NK cells and CTLs attempt to clear mutated tissue cells, and deactivate upon successfully carrying out their cytotoxic function. Tregs inhibit cytotoxic activity. Tregs also release TGF-$\beta$ which promotes the recruitment of Tregs, as well as increasing the rate of EMT.
\par
The inflammation cycling scheme for a typical {\it in silico} patient consists of alternating high and low regimes (Fig. \ref{fig:ModelIntro}B); the inflammation schemes modeled will be discussed in detail in the next section. For this patient, after the warmup period, cell mutations are observed at a rate low enough that they are cleared by cytotoxic cells before being able to establish a tumor. However, after approximately 1000 cell cycles (750 days), one mutated cell exhibits clonal growth, and a tumor is established. At this point, large numbers of cells from the adaptive immune system (CTLs and Tregs) are being heavily recruited, and a peak in TGF-$\beta$ expression is observed. The number of mesenchymal cells increases initially with increasing TGF-$\beta$, but decreases after the clone takes over the tissue. This shift back towards an epithelial tissue phenotype occurs even though the expression of TGF-$\beta$ remains high.
After 841 cell cycles, the proportion of mutated cells reaches 50\%: this is modeled as the threshold defining tumorigenesis. Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days. 
After cancer incidence, the number of mutated cells continues to grow rapidly and soon makes up 100\% of the cell population. A peak in mesenchymal phenotypes is observed following cancer incidence, before most transition back to an epithelial state.
\par 
Considering the immune system dynamics, we see that the NK population is approximately constant, while the adaptive populations (CTL and Treg cells) grow quickly and dwarf NK cells in number following the accumulation of oncogenic mutations.
The adaptive immune populations also appear to exhibit oscillatory behavior, however note that this is not due to intrinsic dynamics but rather due to the inflammation scheme that the patient is undergoing: alternating between 30 cell cycles of high inflammation and 60 cell cycles of low inflammation. Within each of these periods, adaptive immune populations increase or decrease rapidly in accordance with the inflammation state.
\par 
The {\em in silico} patient described here is given for the purpose of illustrating features of the model. Given the multiple sources of noise in the model, in order to quantify patient dynamics and cancer-free survival rates, below we will simulate large cohorts of patients. In Fig. \ref{fig:ModelIntro}C we simulate the survival curve for a cohort of $500$ patients: we see that all the patients survive for approximately 100 cell cycles (75 days). Subsequently, in the approximate range of $T= [100,300]$, the cancer onset rate is roughly constant, and after $600$ cell cycles, no patients remain cancer free.

\subsection{Identification of key regulatory parameters via Morris global sensitivity analysis}\label{SensAnalysis}
To assess the sensitivity of the model parameters and thus identify those that are most important in determining cancer-free survival times, we applied Morris one-step-at-a-time (OAT) sensitivity analysis (see Methods) to the model.
The results of Morris OAT on the 31 model parameters are shown in Figure \ref{fig:MOAT}: a subset of parameters demonstrate much higher levels of sensitivity than others.
The two most influential according to this analysis are the death rate and the recruitment rate of Treg cells, this is most likely due to the dual roles Treg cells play in both suppressing the cytotoxic effects of other immune cells and secreting TGF-$\beta$, which drives EMT.
This ties Treg cells to all three components of the model.
Since we seek to separate the effects of different model components, we do not choose the parameters influencing Treg cells for detailed analysis below.
\par
Many parameters in the model change depending on the inflammation state.
For the sake of a naming convention, any parameter which includes ``low'' in the name represents the parameter value when inflammation is low.
For when the inflammation is high, these ``low'' parameters get scaled by parameters which include ``up'' in the name.
The ILD and IHD parameters determine the duration of low and high inflammation, respectively.
A cell with mutated pathways will be affected by the $\Delta_A$, $\Delta_P$, and $\Delta_{\text{IE}}$ parameters, depending on whether the apoptosis, proliferation, or immune evasion pathways are mutated.
The mesenchymal cells are influenced by the MGA and MIE parameters.
Finally, $\sigma$, $k_\text{EMT}$, $\tau_\text{max}$, and $K_3$ all control how cells transition between the epithelial and mesenchymal states.

@@

\begin{figure}
\center
{\includegraphics[width=0.85\textwidth]{Figure3/Figure3.pdf}}
\caption{Effects of mesencymal cell properites on the time to cancer. Trajectories of one patient per cohort from warmup period ([-1000,0]) to 2000 cell cycles, for $\Delta_\text{MIE}$ (A); $\Delta_\text{MGA}$ (B); and $\tau_\text{Treg}$ (C). 
D. Survival curves corresponding to changes in MIE (A) for a patient cohort of size 1000; shaded region represents the 95\% confidence interval for the evaluated function. 
E. Survival curves for patient cohort corresponding to changes in MGA (B).
F. Survival curves for patient cohort corresponding to changes in Treg TGF-$\beta$ (C). 
%% \color{red} Cox regression or KM test statistics to be computed to demonstrate the significance of this claim.
}
\label{fig:FirstSurvivalCurves}
\end{figure}

When a cell transitions from an epithelial to a mesenchymal state, two phenotypic cell characteristics change: mesenchymal immune evasion (MIE) and mesenchymal growth arrest (MGA).
Both parameters are defined as proportions, and lie in $[0,1]$, where higher values indicate more mesenchymal-like properties.
The parameter $\Delta_\text{MIE}$ is the proportional reduction in the probability that a mutated mesenchymal cell will be subject to immune clearance in a given cycle.
The parameter $\Delta_\text{MGA}$ is the proportional reduction in the probability that a mesenchymal cell will proliferate in a given cell cycle, thus is equivalent (strictly in terms of its impact on the cell cycle) to an increased proportion of time spent resting in the $G_0$ phase.
The cytokine TGF-$\beta$ is also involved in the EMT process (by increasing the probability of EMT). All three of these parameters were found to have large effects on the model by the Morris OAT analysis presented in Section \ref{SensAnalysis}.
\par 
As MIE is increased, the Time to Cancer decreases (Fig. \ref{fig:FirstSurvivalCurves}A) under all sets of parameters studied: as this subpopulation of potentially mutated cells becomes more resistant to immune clearance, the tumor as a whole grows more resilient and thus will grow faster.
These results are summarized by Figure \ref{fig:FirstSurvivalCurves}A.

Second, as MGA increases, Time to Cancer increases, i.e. lower proliferation rates for mesenchymal cells slow down cancer growth (Figure \ref{fig:FirstSurvivalCurves}B).
This is not immediately intuitive, since the decreased proliferation rate affects both mutated and non-mutated tissue cells.

@@

We explored the effects of varying the inflammation state of the patient on cancer-free survival, to investigate competing interactions within the TME and their effect on EMT. Patient cohorts were simulated under different inflammation regimes: permanently low inflammation; permanently high inflammation; or variable inflammation. For patients drawn from cohorts in a permanently high inflammatory state, the relationship between mesenchymal parameters and the Time to Cancer is monotonic, i.e. increasing either MIE (Fig. \ref{fig:VaryINFL_and_MesPars}A-B) or MGA (Fig. \ref{fig:VaryINFL_and_MesPars}C-D) decreases the Time to Cancer.
However, under regimes with either temporary or permanent periods of low inflammation, different relationships emerge: a local maximum for the Time to Cancer is found with respect to MGA ($\Delta_\text{MGA}= 0.3$). This is seen both for intermittent or low inflammation (Fig. \ref{fig:VaryINFL_and_MesPars}D). 
\par
These striking differences in the mean Time to Cancer -- extended by up to one year by fine-grained optimization of mesenchymal growth arrest -- have clear therapeutic implications. 
Our model predicts that reducing the inflammatory state of the TME in three weeks out of every nine would be provide means to prolong cancer-free survival in combination with mesenchymal cell therapies to make the control of mesenchymal proliferation an effective anti-cancer strategy.
\par
In contrast, when MIE is varied for different inflammation cycling schemes, regardless of the cycling scheme, increasing the MIE leads to decreases in the Time to Cancer (worse survival probabilities), although in the case where the inflammatory state is permanently high, MIE has little effect on the Time to Cancer. Thus under any inflammation regime with periods of low inflammation, any reductions in the rate of mesenchymal immune evasion will lead to improved patient outcomes.
